---
title: "R Notebook"
output: html_notebook
---

```{r}
library(tidyverse)
library(lubridate)
library(readxl)
install.packages("tidylog")
library(tidylog)
```

## Counting and summarizing

1. `count()`

2. Creating variables in `count()`

3. `add_count()`

## Wrangling data 

4. `str_pad()`

5. `replace_na()`

6. `lag()`

## tidy data

7. `crossing()`

8. `separate()`

9. **collapse**

10. **tidylog**

So much of data science is counting, and sometimes dividing

# 1. `count()`

## Regular summarizing 

```{r mlb, echo=T}
mlb <- read_csv("data/mlb.csv")
glimpse(mlb)
```

```{r, echo=T}
mlb %>% 
  group_by(team) %>% 
  summarize(total=n())
```



## Using count()

```{r mlb2, echo=T}
mlb <- read_csv("data/mlb.csv")
glimpse(mlb)
```

```{r, echo=T}
mlb %>% 
  count(team, league)

saved <- mlb %>% 
  select(team, league) %>% 
  unique()

saved_count <- saved %>% 
  count(team)
```


## Using count() and names

```{r mlb3, echo=T}
mlb <- read_csv("data/mlb.csv")
glimpse(mlb)
```

```{r, echo=T}
mlb %>% 
  count(team, 
        name="total")
```




## Using count() and more


```{r mlb4, echo=T}
mlb <- read_csv("data/mlb.csv")
glimpse(mlb)
```


```{r, echo=T}
mlb %>% 
  count(team, 
        sort=TRUE,
        wt=salary,
        name="total_salary")
```



# 2. Creating new variables in a `group_by()` or `count()`

## Bring in Disney data

```{r echo=T}
disney <- read_csv("data/disney_movies_total_gross.csv")

glimpse(disney)
```

## count() again

```{r, echo=T}
disney %>% 
  mutate(year=year(mdy(release_date))) %>% 
  mutate(month=month(mdy(release_date))) %>% 
  count(year, month) %>% 
  arrange(month, year)
```


## count() again but better

```{r, echo=T}
disney %>% 
  mutate(year=year(mdy(release_date))) %>% 
  count(decade = 10 * (year %/% 10))
```

## excellent for quicker viz

```{r, echo=T}
disney %>% 
  mutate(year=year(mdy(release_date))) %>% 
  count(year) %>% 
  ggplot(aes(x=year, y=n)) +
  geom_col() 
```

## excellent for quicker viz

```{r, echo=T}
disney %>% 
  mutate(year=year(mdy(release_date))) %>% 
  count(decade = 10 * (year %/% 10)) %>% 
  ggplot(aes(x=decade, y=n)) +
  geom_col() 
```


# 3.add_count()

## add_count()

```{r, echo=T}
osha <- read_csv("data/osha.csv")

glimpse(osha)
```

## add_count()

Normally:

```{r, echo=T}
osha %>% 
  mutate(city=str_to_title(city)) %>% 
  group_by(city, state) %>% 
  summarize(total=n()) %>% 
  filter(total>100)
```

## add_count()

Normally:

```{r, echo=T}
osha %>% 
  mutate(city=str_to_title(city)) %>% 
  group_by(city, state) %>% 
  summarize(total=n()) %>% 
  group_by(state) %>% 
  mutate(state_count=sum(total)) %>% 
  filter(total>100)
```

## add_count() 

```{r echo=T}
osha %>% 
  mutate(city=str_to_title(city)) %>% 
  count(city, state) %>% 
  add_count(state, wt=n) %>% 
  filter(n>100)
```

## add_count() 

```{r echo=T}
osha %>% 
  mutate(city=str_to_title(city)) %>% 
  count(city, state, name="city_count") %>% 
  add_count(state, wt=city_count, name="state_count") %>% 
  filter(city_count>100)
```

# 4. str_pad() when you need leading zeroes

## str_pad()

```{r echo=T}
poverty <- read_csv("data/poverty_original.csv") %>% 
  mutate(bad_fips=as.numeric(GEOID))

glimpse(poverty)
```

## str_pad() 

```{r echo=T}
poverty <- read_csv("data/poverty_original.csv") %>% 
  mutate(bad_fips=as.numeric(GEOID)) %>% 
  mutate(good_fips=str_pad(bad_fips, 5, side="left", "0"))

glimpse(poverty)
```

# 5. replace_na()

## replace_na()

```{r echo=T}
transactions <- read_csv("data/transactions.csv")

glimpse(transactions)
```

## replace_na()

```{r  echo=T}
transactions <- transactions %>% 
  mutate(employer = replace_na(employer, 0))

glimpse(transactions)
```

## replace_na()

```{r echo=T}
transactions <- transactions %>% 
  mutate(employer = replace_na(employer, 0),
         occupation =replace_na(occupation, 0),
         memo_cd = replace_na(memo_cd, ""))

glimpse(transactions)
```

# 6. lag

## lag()

```{r echo=T}
disney %>% 
  mutate(year=year(mdy(release_date))) %>% 
  count(year)
```

## lag()

```{r echo=T}
disney %>% 
  mutate(year=year(mdy(release_date))) %>% 
  count(decade = 10 * (year %/% 10)) %>% 
  mutate(prior_decade=lag(n, 1))
```

## lag()

```{r echo=T}
disney %>% 
  mutate(year=year(mdy(release_date))) %>% 
  count(decade = 10 * (year %/% 10)) %>% 
  mutate(prior_decade=lag(n, 1)) %>% 
  mutate(change=round(n-prior_decade)/prior_decade*100,1)
```

# 7. crossing()

## Making a dummy set with crossing()

```{r echo=T}
dummy_df <- crossing(month=1:12,
                     year=2019:2022)

dummy_df
```

## Making a dummy set with crossing()

```{r echo=T}
state.abb
```

## Making a dummy set with crossing()

```{r echo=T}
dummy_df <- crossing(month=1:12,
                     year=2019:2022,
                     state=state.abb)

dummy_df
```

# 8. separate()

## separate()

```{r echo=T}
candidates <- read_csv("data/candidates.csv")

glimpse(candidates)
```

## separate()

```{r echo=T}
candidates_new <- candidates %>% 
  mutate(cand_original=cand_name) %>% 
  separate(cand_name, sep=", ", into=c("last", "first"))

glimpse(candidates_new)
```

# 9. collapse

## collapse in paste0()

```{r echo=TRUE}
sal <- read_excel("data/Salaries.xlsx", sheet=1)

glimpse(sal)
```

## collapse in paste0()

```{r echo = TRUE}
sal %>% group_by(department) %>%
  summarize(total_budget=sum(total_comp, na.rm=T),
            titles=paste0(job_title, collapse=", "))
```

## collapse in paste0()

```{r echo=TRUE}
sal %>% group_by(department) %>%
  summarize(total_budget=sum(total_comp, na.rm=T),
            titles=paste0(unique(job_title), collapse=", "))
```

# 10. tidylog

## tidylog

```{r echo=TRUE}
library(tidylog)

transactions <- read_csv("data/transactions.csv", col_types="ccccccccccccccTnccccccc")
candidates <- read_csv("data/candidates.csv", col_types = cols(.default = "c"))
```

## tidylog

```{r echo=TRUE}
inner_join(transactions, candidates, by = "cand_id")
```

## tidylog

```{r echo=T}
left_join(transactions, candidates, by = "cand_id")
```

